\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-02 papers/paper1/ms.tex
some words about Cholesky factorization

\section{Implementation \& performance}

@@

>>>>

\section{Semi-separable matrix factorization}

\subsection{Cholesky factorization}

The relationship between the \celeriteterm\ model and semi-separable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
For the \celeriteterm\ model described in the previous section, the rank of
the semi-separable matrix is $R = 2\,J$ and the components of the relevant
matrices are
\begin{eqnarray}
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j+1} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j+1} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{0,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

\subsection{Solving}

After computing the Cholesky factorization of the matrix $K$, we can apply
the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \log \det K = \sum_{n=1}^N \log D_{n,n}
\end{eqnarray}
where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& f_{n-1,j} + X_{n-1,j}\,z_{n-1} \nonumber\\
    z_n &=& y_n - \sum_{j=1}^{2\,J} U_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{b}$
\begin{eqnarray}
    f_{n,j} &=& f_{n+1,j} + U_{n+1,j}\,b_{n+1} \nonumber\\
    b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} X_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

\subsection{Pre-conditioning}

@@

\section{An alternative solver using the extended matrix formalism}
